---
title: "March Madness Model 2026"
author: "Vanhi Chava"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r, output = FALSE}
library(dplyr)
library(tidyverse)
```


```{r}
ratings_538 <- read.csv("datasets/538_right_ratings.csv")
ap_poll <- read.csv("datasets/AP_Poll.csv")
bart_neut <- read.csv("datasets/BartNeutral.csv")
ken_pom_bart <- read.csv("datasets/KPBart.csv")
tourney_past_games <- read.csv("datasets/Tournament Matchups.csv")
team_rankings <- read.csv("datasets/teamranks.csv")
```


# Potential Features to Use

**Data Link:**

All datasets taken from: https://www.kaggle.com/datasets/nishaanamin/march-madness-data/data

**538 Ratings:**

* only goes to 2023

potential features: POWER.RATING (double, 0 to 100)

**AP Poll:**

* potential features: AP.RANK (1 to 47)

**Barttorvik Neutral:**

* EFG., EFG.D

* TOV. , TOV.D

* ELITE.SOS

* TALENT

* WIN. (Win Percentage)

* BARTHAG (team strength rating, adjusted offensive and defensive efficiency), (close to 1 elite team, 0.5 average, 0.4 and below is weak)

* BADJ.O (team offensive impact per 100 possession, based on box score, + helps offense and - hurts offense)

* BADJ.D (team defensive impact per 100 possession points prevent, negative is good)

* ORB., DRB.

* EXP (experience of players)


**KenPom Barttorvik:** (in addition to neutral)

* KADJ.EM --> adjusted efficiency margin, represents number of points team would outscore an avg. opponent by over 100 possessions (almost always best single-variable predictor)

* KADJ.O & KADJ.D --> KenPom's version of adjusted offensive and defensive efficiency

* KADJ.T (adjusted tempo) --> possessions per 40 minutes, adjusted for opponents (look for tempo mismatches, faster teams will win)

* X3PT. --> 3 point percentage

* X2PT. --> 2 point percentage



**Tournament Matchups:**

Holds all previous tournament matchups and who won.

**Team Rankings:**

* SOS.RATING

* V.1.25.WINS & V.1.25.LOSS --> perfromance against the elite teams

* V.26.50 WINS 

* CONSISTENCY.RANK --> team's performance from game to game ranked

* LUCK.RATING --> deviation between team's actual win-loss record and expected record based on efficiency (high-luck teams will not have as much luck in the tournament)



# Data Cleaning + Preparation

Picking the main predictors + creating a master dataset:

```{r}
# pick the predictors from each dataset, also keep YEAR and TEAM
ratings_538_small <- ratings_538 %>%
  select(YEAR, TEAM, POWER.RATING)

ap_small <- ap_poll %>%
  select(YEAR, TEAM, AP.RANK)

bart_small <- bart_neut %>%
  select(YEAR, TEAM, EFG., EFG.D, TOV., TOV.D, ELITE.SOS, TALENT, WIN., BARTHAG, BADJ.O, BADJ.D, OREB., DREB., EXP)

ken_pom_small <- ken_pom_bart %>%
  select(YEAR, CONF, TEAM, KADJ.EM, KADJ.O, KADJ.D, KADJ.T, X2PT., X3PT.)

team_rankings_small <- team_rankings %>%
  select(YEAR, TEAM, SOS.RATING, V.1.25.WINS, V.1.25.LOSS, CONSISTENCY.RANK, LUCK.RATING)

# join by YEAR and TEAM
master_df <- ken_pom_small %>%
  left_join(bart_small, by = c("YEAR", "TEAM")) %>%
  left_join(team_rankings_small, by = c("YEAR", "TEAM")) %>%
  left_join(ratings_538_small, by = c("YEAR", "TEAM")) %>%
  left_join(ap_small, by = c("YEAR", "TEAM")) 

# team name changes
master_df <- master_df %>%
  mutate(TEAM = case_when(
    TEAM == "Iowa St."             ~ "Iowa St",
    TEAM == "McNeese St."          ~ "McNeese St",
    TEAM == "Michigan St."         ~ "Michigan St",
    TEAM == "North Carolina St."   ~ "NC State",
    TEAM == "Ohio St."             ~ "Ohio St",
    TEAM == "St. John's"           ~ "St John's",
    TEAM == "Saint Louis"           ~ "St Louis",
    TEAM == "Saint Mary's"         ~ "St Mary's CA",
    TEAM == "Utah St."             ~ "Utah St",
    TEAM == "Kennesaw St."         ~ "Kennesaw",
    TEAM == "North Dakota St."     ~ "N Dakota St",
    TEAM == "Tennessee St."        ~ "Tennessee St",
    TEAM == "Wright St."           ~ "Wright St",
    TEAM == "Queens"               ~ "Queens NC", 
    TEAM == "Prairie View A&M"     ~ "Prairie View",
    TRUE ~ TEAM
  ))

master_df <- master_df %>%
  distinct(YEAR, TEAM, .keep_all = TRUE)
```

Prepping the Tournament Matchups dataset so each game is one row + Cleaning:

```{r}
tourney_results <- tourney_past_games %>%
  filter(!is.na(SCORE) & SCORE > 0) %>%
  group_by(YEAR, CURRENT.ROUND, id = ceiling(row_number() / 2)) %>%
  summarize(
    WINNER = TEAM[which.max(SCORE)],
    LOSER = TEAM[which.min(SCORE)],
    SEED.W = SEED[which.max(SCORE)],
    SEED.L = SEED[which.min(SCORE)],
    SCORE.W = max(SCORE),
    SCORE.L = min(SCORE),
    ROUND = first(ROUND),
    .groups = "drop") %>%
  mutate(SCORE.DIFF = SCORE.W - SCORE.L) %>%
  select(YEAR, WINNER, LOSER, SEED.W, SEED.L, ROUND, CURRENT.ROUND, SCORE.W, SCORE.L, SCORE.DIFF) %>%
  arrange(desc(YEAR))
```


# Training Dataset

```{r}
# non-breaking spaces with regular spaces replacement
master_df$TEAM <- gsub("\u00A0", " ", master_df$TEAM)
tourney_results$WINNER <- gsub("\u00A0", " ", tourney_results$WINNER)
tourney_results$LOSER <- gsub("\u00A0", " ", tourney_results$LOSER)

# joining stats for both teams by joining "tourney_results" and "master_df"
winner_stats <- master_df %>%
  rename_all(~paste0(., ".W")) 
loser_stats <- master_df %>%
  rename_all(~paste0(., ".L")) 

training_data <- tourney_results %>% 
  left_join(winner_stats, by = c("YEAR" = "YEAR.W", "WINNER" = "TEAM.W")) %>%
  left_join(loser_stats, by = c("YEAR" = "YEAR.L", "LOSER" = "TEAM.L"))

# adding difference feature
training_data <- training_data %>%
  mutate(across(c(KADJ.EM.W, KADJ.EM.L, BARTHAG.W, BARTHAG.L, SEED.W, SEED.L), as.numeric)) %>%
  mutate(KADJ.EM_DIFF = KADJ.EM.W - KADJ.EM.L,
   BARTHAG_DIFF = BARTHAG.W - BARTHAG.L,
   SEED_DIFF = SEED.W - SEED.L,
   Result = 1)

final_training_data <- training_data %>%
  filter(!is.na(KADJ.EM_DIFF))
```



# Model Fitting

```{r}
# creating a model ready data.frame
wins <- final_training_data %>%
  mutate(Result = 1)

losses <- final_training_data %>%
  mutate(
    KADJ.EM_DIFF = -KADJ.EM_DIFF,
    LUCK.RATING.W_wait = LUCK.RATING.W,
    LUCK.RATING.W = LUCK.RATING.L,
    LUCK.RATING.L = LUCK.RATING.W_wait,
    CONSISTENCY.RANK.W_wait = CONSISTENCY.RANK.W,
    CONSISTENCY.RANK.W = CONSISTENCY.RANK.L,
    CONSISTENCY.RANK.L = CONSISTENCY.RANK.W_wait,
    Result = 0) %>%
  select(-LUCK.RATING.W_wait, -CONSISTENCY.RANK.W_wait)

model_df <- bind_rows(wins, losses)

model_df <- model_df %>%
    mutate(LUCK_DIFF = LUCK.RATING.W - LUCK.RATING.L,
         CONSISTENCY_DIFF = CONSISTENCY.RANK.W - CONSISTENCY.RANK.L)
```

Model:

```{r}
# split into training and test data
ready_training_data <- model_df %>%
  filter(YEAR < 2024)

testing_data <- model_df %>%
  filter(YEAR >= 2024)

mm_model_1 <- glm(Result ~ KADJ.EM_DIFF + LUCK_DIFF + CONSISTENCY_DIFF, 
                  data = ready_training_data,
                  family = "binomial")
summary(mm_model_1)
```

## Checking Brier Score + Most Games Correct

```{r}
# test set probabilities
testing_data$probability <- predict(mm_model_1, testing_data, type = "response")

# brier score calculation
brier_score <- mean((testing_data$probability - testing_data$Result)^2)
brier_score

# most games correct
testing_data <- testing_data %>%
  mutate(prediction = ifelse(probability > 0.5, 1, 0),
         correct_games = ifelse(prediction == Result, 1, 0))
mean(testing_data$correct_games)
```



# Submission Formatting

```{r}
potential_matchups <- read.csv("2026_Potential_Matchups.csv")

stats_2026 <- master_df %>% filter(YEAR == 2026)

# join stats for higher seed
submission_df <- potential_matchups %>%
  left_join(stats_2026, by = c("HigherSeed" = "TEAM")) %>%
  rename(
    KADJ.EM.W = KADJ.EM,
    LUCK.RATING.W = LUCK.RATING,
    CONSISTENCY.RANK.W = CONSISTENCY.RANK
  )

# join stats for lower seed
submission_df <- submission_df %>%
  left_join(stats_2026, by = c("LowerSeed" = "TEAM")) %>%
  rename(
    KADJ.EM.L = KADJ.EM,
    LUCK.RATING.L = LUCK.RATING,
    CONSISTENCY.RANK.L = CONSISTENCY.RANK
  )

# features I need
submission_df <- submission_df %>%
  mutate(KADJ.EM_DIFF = KADJ.EM.W - KADJ.EM.L,
         LUCK_DIFF = LUCK.RATING.W - LUCK.RATING.L,
         CONSISTENCY_DIFF = CONSISTENCY.RANK.W - CONSISTENCY.RANK.L)

close_submission_data <- submission_df %>%
  select(HigherSeed, HigherSeedID, HigherSeedNum, LowerSeed, LowerSeedID, LowerSeedNum, Predictions, KADJ.EM_DIFF, LUCK_DIFF, CONSISTENCY_DIFF)

close_submission_data$Predictions <- predict(mm_model_1, close_submission_data, type = "response")

final_submission_data <- close_submission_data %>%
  select(HigherSeed, HigherSeedID, HigherSeedNum, LowerSeed, LowerSeedID, LowerSeedNum, Predictions)
  
write.csv(final_submission_data, "Vanhi_Chava_SAC_Submission.csv", row.names = FALSE)
```



