---
title: "Nba2025"
output: html_document
date: "2025-06-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

# Initial Data Import and Manipulation

```{r}
nba <- read.csv("C:/Users/VanhiChava/Downloads/nba2025.csv")
head(nba)
```


# Question Exploration

## What are the top 3 teams that had the highest mean PPG average amongst all untraded players?
```{r}
PPG_nba <- nba %>% 
  filter(CUR == "Yes") %>%
  group_by(TEAM) %>%
  summarize(sumPPG = sum(PpG),
            total_players = n(),
            meanPPG = sumPPG/total_players)
ggplot(PPG_nba, aes(x = meanPPG, y = TEAM, fill = TEAM)) + 
    geom_col() + 
    labs(title = "NBA Teams' Mean PPG per Player")
```
```{r}
PPG_nba %>%
  arrange(meanPPG) %>%
  slice_max(meanPPG, n = 3)
```

> The teams with the highest mean PPG among untraded players are: New Orleans Pelicans, Atlanta Hawks, and the Dallas Mavericks. 

## What are the top 5 oldest and youngest NBA teams prior to the start of the 2025 NBA season?

```{r}
age_nba <- nba %>%
  group_by(TEAM) %>%
  summarize(meanAge = sum(AGE) / n())

ggplot(age_nba, aes(x = meanAge, y = TEAM, color = TEAM)) + 
  geom_point(size = 3.0, shape = 10) + 
  theme_classic()
```

```{r}
age_nba %>%
  arrange(meanAge) %>%
  slice_max(meanAge, n = 5)
```


> The top 5 oldest teams in the NBA are the Phoenix Suns, the Los Angeles Clippers, the Golden State Warriors, the Miami Heat, and the Boston Celtics.


```{r}
age_nba %>%
  arrange(meanAge) %>%
  slice_min(meanAge, n = 5)
```

> The top 5 youngest teams in the NBA are the Portland Trail Blazers, the Brooklyn Nets, the Oklahoma City Thunder, the Utah Jazz, and the Memphis Grizzlies.

NEXT TIMES: create density plots of ages for the oldest and youngest teams. 


## What does the age distribution of players for the NBA teams with the youngest and oldest overall mean age?

Youngest NBA Team: 

```{r}
young_nba <- age_nba %>%
  arrange(meanAge) %>%
  slice_min(meanAge, n = 1) 
young_team <- young_nba$TEAM[1]

nba %>%
  filter(TEAM == young_team) %>%
  select(NAME, AGE) %>%
  ggplot(aes(x = AGE)) + 
    geom_density(fill = "dodgerblue", alpha = 0.5) + 
    labs(title = "Age Distribution Plot of Youngest NBA Team")
  
```
Oldest NBA Team: 

```{r}
old_nba <- age_nba %>%
  arrange(meanAge) %>%
  slice_max(meanAge, n = 1) 
old_team <- old_nba$TEAM[1]

nba %>%
  filter(TEAM == old_team) %>%
  select(NAME, AGE) %>%
  ggplot(aes(x = AGE)) + 
    geom_density(fill = "red", alpha = 0.5) + 
    labs(title = "Age Distribution Plot of Oldest NBA Team")
  
```
In order to have a more direct comparison, we will overlap the plots.
```{r}
nba %>%
  filter(TEAM == old_team | TEAM == young_team) %>%
  ggplot(aes(x = AGE, fill = TEAM, alpha = 0.2)) + 
    geom_density()
  
```





















